Fractional Types

### 2013-02-23 frac.tex
some notes

In the computer science community, the theory and practice of type
isomorphisms is well-established. Originally, such type isomorphisms were
motivated by the pragmatic concern of searching large libraries of functions
by providing one of the many possible isomorphic types for the desired
function~\cite{Rittri:1989:UTS:99370.99384}. More recently, type isomorphisms
have taken a more central role as \emph{the} fundamental computational
mechanism from which more conventional, i.e., irreversible computation, is
derived. In more detail, in our own previous
work~\cite{James:2012:IE:2103656.2103667,rc2011,rc2012} we started with the
notion of type isomorphism as the ``right'' starting point for a theory of
computation. The central technical contribution was the development of a
family of programming languages, {{Pi}} with various superscripts, in which
every computation is an isomorphism. Other major results are that all such
computations preserve the information-theoretic entropy, that the languages
are logically reversible, and that they have models in various flavors of
symmetric monoidal categories (which are the backbone of models for linear
logic and quantum computation). We have also invested a significant amount of
time in using {{Pi}} as an actual programming language and established
several idioms for programming large classes of interesting (recursive)
programs in {{Pi}}.

A major open problem remains, however: a higher-order extension of
{{Pi}}. This extension is of fundamental importance in all the originating
research areas. In physics, it allows a process or observer to be treated as
``data'' that can be transformed by higher-level processes or observed by
meta-level observers. In mathematics and logic, it allows the equivalence
between different proofs of type $\texttt{Id}_A(a,b)$ to itself be expressed
as an isomorphism (of a higher type)
$\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
higher-order types allow code to abstract over other code fragments as well
as the manipulation of code as data and data as code.

>>>>

\roshan{The above para is too technical in a weird way. Its unclear
  what it contributes. If one wants to make a point of computational
  witnesses of isomorphisms its easier to explain and cite in terms of
  Andreas Blass and Fiore -- and more generally by Barry Mazur. I say
  this also because it seems that HOTT is more concerned with
  topological structure of equalities, even though they tend to start
  their literature with some talk of intentional specifications.}

In the computer science community, the theory and practice of type
isomorphisms is well-established. Originally, such type isomorphisms were
motivated by the pragmatic concern of searching large libraries of functions
by providing one of the many possible isomorphic types for the desired
function~\cite{Rittri:1989:UTS:99370.99384}. More recently, type isomorphisms
have taken a more central role as \emph{the} fundamental computational
mechanism from which more conventional, i.e., irreversible computation, is
derived. In our own previous
work~\cite{James:2012:IE:2103656.2103667,rc2011,rc2012} we started with the
notion of type isomorphism as the ``right'' starting point for a theory of
computation. The central technical contribution was the development of a
family of programming languages, {{Pi}} with various superscripts, in which
every computation is an isomorphism. Other major results are that all such
computations preserve the information-theoretic entropy, that the languages
are logically reversible, and that they have models in various flavors of
symmetric monoidal categories (which are the backbone of models for linear
logic and quantum computation). 
%% We have also invested a significant amount of time in using {{Pi}} as
%% an actual programming language and established several idioms for
%% programming large classes of interesting (recursive) programs in
%% {{Pi}}.

A major open problem remains, however: a higher-order extension of
{{Pi}}. This extension is of fundamental importance in all the originating
research areas. 
%% In physics, it allows a process or observer to be treated as
%% ``data'' that can be transformed by higher-level processes or observed by
%% meta-level observers. 
In physics, it allows for quantum states to be viewed as processes and
processes to be viewed as states, such as with the Choi-Jamiolkowski
isomorphism \cite{choi1975completely,jamiolkowski1972linear}.  In
mathematics and logic, it allows the equivalence between different
proofs of type $\texttt{Id}_A(a,b)$ to itself be expressed as an
isomorphism (of a higher type)
$\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
higher-order types allow code to abstract over other code fragments as
well as the manipulation of code as data and data as code.

\roshan{Again, HOTT seems jarring. Saying something about logic in
  this way however seems useful. Maybe one can use one of the examples
  of equivalence between proofs that Baez's Rosetta stone uses. }
